<div style="text-align: center;">
  <h1>Company Bankruptcy Prediction</h1>
  <h4>Angela Wei, Aditya Desai, Andrew Theiss, and Samuel Gadkar</h4>
  <h4>December 14th, 2025</h4>
</div>

<br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
# Import Libraries
library(tidyverse)
library(rpart)
library(rpart.plot)
library(caret)
library(pROC)
library(class)
library(pls)
library(glmnet)
library(randomForest)
library(readr)
library(ggplot2)
library(dplyr)
library(rlang)
library(gridExtra)
```

```{r, include = FALSE}
# Set See and Import Data
set.seed(4620)
train <- read.csv("../data/train.csv")
test <- read.csv("../data/test.csv")
```

<center>
## Introduction

???

## Exploratory Data Analysis

???

### Data Problems
```{r}

```

???

### Variable Types
```{r}

```

???

### Missing Data
```{r}

```

???

### Collinearity Issues
```{r}

```

???

### Potential Important Predictors
```{r}

```

???

### Key Figures and Numerical Summaries
```{r}

```

???

### Potential Models
```{r}

```

???

## Model Analysis

### Investigated Models

???

#### Ridge Regression
```{r, include = FALSE}
# Setup Ridge Data
train_ridge <- train
test_ridge <- test

y_train_ridge <- as.numeric(as.character(train_ridge$Bankrupt.))
y_test_ridge <- as.numeric(as.character(test_ridge$Bankrupt.))

# Cleanup Columns
predictor_cols_ridge <- setdiff(colnames(train_ridge), c("Bankrupt.", "Unnamed..0"))

X_train_ridge <- as.matrix(train_ridge[, predictor_cols_ridge])
X_test_ridge <- as.matrix(test_ridge[, predictor_cols_ridge])

# Balanced Class Weights
class_counts <- table(train_ridge$Bankrupt.)
minority_weight <- class_counts["0"] / class_counts["1"]

ridge_weights <- ifelse(train_ridge$Bankrupt. == "1", minority_weight, 1)

# Fit Ridge Logistic Regression
ridge_cv <- cv.glmnet(
  x = X_train_ridge,
  y = y_train_ridge,
  alpha = 0,  
  family = "binomial",
  standardize = TRUE,
  weights = ridge_weights,
  type.measure = "deviance"
)

best_lambda_ridge <- ridge_cv$lambda.min
best_lambda_ridge

# Fit Final Ridge Model
ridge_fit <- glmnet(
  x = X_train_ridge,
  y = y_train_ridge,
  alpha = 0,
  family = "binomial",
  lambda = best_lambda_ridge,
  standardize = TRUE,
  weights = ridge_weights
)

# Predict on Test Data
ridge_prob <- predict(ridge_fit, newx = X_test_ridge, type = "response")

ridge_pred <- ifelse(ridge_prob > 0.5, "1", "0")
ridge_pred <- factor(ridge_pred, levels = c("0", "1"))

# Evaluate Ridge Performance
y_test_factor <- factor(test_ridge$Bankrupt., levels = c("0", "1"))
conf_mat_ridge <- confusionMatrix(ridge_pred, y_test_factor)
conf_mat_ridge

accuracy_ridge <- conf_mat_ridge$overall["Accuracy"]
sensitivity_ridge <- conf_mat_ridge$byClass["Sensitivity"]
specificity_ridge <- conf_mat_ridge$byClass["Specificity"]

accuracy_ridge
sensitivity_ridge
specificity_ridge
```

???

#### LASSO
```{r, include = FALSE}
# Setup LASSO Data
train_lasso <- train
test_lasso  <- test

y_train_lasso <- as.numeric(as.character(train_lasso$Bankrupt.))
y_test_lasso  <- as.numeric(as.character(test_lasso$Bankrupt.))

predictor_cols_lasso <- setdiff(colnames(train_lasso), c("Bankrupt.", "Unnamed..0"))

X_train_lasso <- as.matrix(train_lasso[, predictor_cols_lasso])
X_test_lasso  <- as.matrix(test_lasso[, predictor_cols_lasso])

# Balanced Class Weights
class_counts <- table(train_lasso$Bankrupt.)
minority_weight <- class_counts["0"] / class_counts["1"]

lasso_weights <- ifelse(train_lasso$Bankrupt. == "1", minority_weight, 1)

# Fit LASSO Logistic Regression
lasso_cv <- cv.glmnet(
  x = X_train_lasso,
  y = y_train_lasso,
  alpha = 1,
  family = "binomial",
  standardize = TRUE,
  weights = lasso_weights, 
  type.measure = "deviance"
)

best_lambda_lasso <- lasso_cv$lambda.min
best_lambda_lasso

# Fit Final Lasso Model
lasso_fit <- glmnet(
  x = X_train_lasso,
  y = y_train_lasso,
  alpha = 1,
  family = "binomial",
  lambda = best_lambda_lasso,
  standardize = TRUE,
  weights = lasso_weights
)

# Predict on Test Data
lasso_prob <- predict(lasso_fit, newx = X_test_lasso, type = "response")

lasso_pred <- ifelse(lasso_prob > 0.5, "1", "0")
lasso_pred <- factor(lasso_pred, levels = c("0", "1"))

# Evaluate LASSO Performance
y_test_factor_lasso <- factor(test_lasso$Bankrupt., levels = c("0", "1"))

conf_mat_lasso <- confusionMatrix(lasso_pred, y_test_factor_lasso)
conf_mat_lasso

accuracy_lasso    <- conf_mat_lasso$overall["Accuracy"]
sensitivity_lasso <- conf_mat_lasso$byClass["Sensitivity"]
specificity_lasso <- conf_mat_lasso$byClass["Specificity"]

accuracy_lasso
sensitivity_lasso
specificity_lasso
```

???

#### Bagging
```{r, include = FALSE}
# Setup Bagging Data
train_bag <- train
test_bag  <- test

train_bag$Bankrupt. <- factor(train_bag$Bankrupt., levels = c("0","1"))
test_bag$Bankrupt.  <- factor(test_bag$Bankrupt., levels = c("0","1"))

predictor_cols <- setdiff(colnames(train_bag), c("Bankrupt.", "Unnamed..0"))

X_train <- train_bag[, predictor_cols]
y_train <- train_bag$Bankrupt.

X_test  <- test_bag[, predictor_cols]
y_test  <- test_bag$Bankrupt.

# Balanced Class Weights
class_counts <- table(y_train)
minority_size <- min(class_counts) 
sampsize_vec <- rep(minority_size, length(class_counts))

# Train Model
bag_model <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  mtry = ncol(X_train), 
  sampsize = sampsize_vec,
  importance = TRUE
)

# Predict on Test Data
bag_prob <- predict(bag_model, X_test, type = "prob")[,2]

bag_pred <- factor(ifelse(bag_prob > 0.5, "1", "0"), levels = c("0","1"))

# Evaluate Model
conf_mat_bag <- confusionMatrix(
  bag_pred,
  y_test,
  positive = "1"
)

conf_mat_bag

accuracy_bag     <- conf_mat_bag$overall["Accuracy"]
sensitivity_bag  <- conf_mat_bag$byClass["Sensitivity"]
specificity_bag  <- conf_mat_bag$byClass["Specificity"]

accuracy_bag
sensitivity_bag
specificity_bag
```

???

#### Random Forest
```{r, include = FALSE}
# Setup Random Forest Data
train_rf <- train
test_rf  <- test

train_rf$Bankrupt. <- factor(train_rf$Bankrupt., levels = c("0","1"))
test_rf$Bankrupt.  <- factor(test_rf$Bankrupt.,  levels = c("0","1"))

predictor_cols <- setdiff(colnames(train_rf), c("Bankrupt.", "Unnamed..0"))

X_train <- train_rf[, predictor_cols]
y_train <- train_rf$Bankrupt.

X_test  <- test_rf[, predictor_cols]
y_test  <- test_rf$Bankrupt.

# Balanced Sampling Setup
class_counts <- table(y_train)
minority_size <- min(class_counts)

sampsize_vec <- rep(minority_size, length(class_counts))

# Fit Random Forest
rf_model <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  mtry = floor(sqrt(ncol(X_train))),
  sampsize = sampsize_vec, 
  importance = TRUE
)

# Predict on Test Set
rf_prob <- predict(rf_model, X_test, type = "prob")[,2]

rf_pred <- factor(ifelse(rf_prob > 0.5, "1", "0"), levels = c("0","1"))

# Confusion Matrix and Metrics
conf_mat_rf <- confusionMatrix(
  rf_pred,
  y_test,
  positive = "1"
)

conf_mat_rf

accuracy_rf     <- conf_mat_rf$overall["Accuracy"]
sensitivity_rf  <- conf_mat_rf$byClass["Sensitivity"]
specificity_rf  <- conf_mat_rf$byClass["Specificity"]

accuracy_rf
sensitivity_rf
specificity_rf
```

???

#### Model Comparison
```{r, warning = FALSE, include = FALSE}
evaluate_model <- function(probabilities, true_labels, threshold = 0.5) {
  
  preds <- factor(ifelse(probabilities > threshold, "1", "0"), levels = c("0","1"))
  
  cm <- confusionMatrix(preds, true_labels, positive = "1")
  
  roc_obj <- roc(true_labels, probabilities)
  auc_val <- as.numeric(auc(roc_obj)) 
  
  tibble(
    Accuracy = as.numeric(cm$overall["Accuracy"]),
    Sensitivity = as.numeric(cm$byClass["Sensitivity"]),
    Specificity = as.numeric(cm$byClass["Specificity"]),
    Balanced_Accuracy = as.numeric((cm$byClass["Sensitivity"] + cm$byClass["Specificity"]) / 2),
    AUC = auc_val
  )
}

ridge_results <- evaluate_model(ridge_prob, y_test_factor)
lasso_results <- evaluate_model(lasso_prob, y_test_factor)
bagging_results <- evaluate_model(bag_prob, y_test_factor)
rf_results <- evaluate_model(rf_prob, y_test_factor)

comparison_table <- bind_rows(
  Ridge = ridge_results,
  LASSO = lasso_results,
  Bagging = bagging_results,
  RandomForest = rf_results,
  .id = "Model"
)
```

???

```{r, warning = FALSE, echo = FALSE}
comparison_table
```


### Description of Final Model
```{r}

```

???

### Model Assumptions
```{r}

```

???

### Model Validation
```{r}

```

???

## Conclusion

???
</center>