---
output:
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

::: {style="text-align: center;"}
<h1>Company Bankruptcy Prediction</h1>

<h4>Angela Wei, Aditya Desai, Andrew Theiss, and Samuel Gadkar</h4>

<h4>December 14th, 2025</h4>
:::

<br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
# Import Libraries
library(tidyverse)
library(rpart)
library(rpart.plot)
library(caret)
library(pROC)
library(class)
library(pls)
library(glmnet)
library(randomForest)
library(readr)
library(ggplot2)
library(dplyr)
library(rlang)
library(gridExtra)
library(knitr)
library(kableExtra)
library(gains)
```

```{r, include = FALSE}
# Set See and Import Data
set.seed(4620)
train <- read.csv("../data/train.csv")
test <- read.csv("../data/test.csv")
```

::: {style="text-align: center;"}
## Introduction
:::

This project analyzes a company bankruptcy dataset. The given
information includes financial statements and other financial metrics,
with the goal of understanding how financial factors play a role in
predicting bankruptcy. The training dataset includes 4,773 observations
and 97 variables. 96 are financial predictors that are numeric. The
response variable, Bankruptcy, is a binary variable. It indicates
whether a firm declared bankruptcy (1), or not (0).

Due to all predictors being numeric, and representing standardized or
normalized financial ratios capturing firm performance, a careful
exploratory analysis is needed to identify data quality issues,
redundancy, and structural challenges that affect modeling.

::: {style="text-align: center;"}
## Exploratory Data Analysis
:::

First we performed exploratory data analysis on our dataset to help
understand the the structure, quality, and key characteristics of the
data. We examine variable types, potential data quality issues such as
extreme values or missing data, relationships between predictors and the
response, and early indications of which variables may be important for
predicting bankruptcy.

::: {style="text-align: center;"}
### Data Problems

A review of summary statistics reveal significant data quality issues.
First, there are many extreme and implausible values. Many predictors
that represent ratios or rates contain extreme values from 10 to 10\^10,
which are not realistic for normalized financial measures. A list is
provided below.
:::

```{r}
num_train <- train[, sapply(train, is.numeric)]

max_vals <- apply(num_train, 2, max, na.rm = TRUE)

bad_vars <- sort(max_vals[max_vals > 1e4], decreasing = TRUE)
bad_vars
```

::: {style="text-align: center;"}
### Variable Types
:::

```{r}
names(train[,-1])

```

Again the dataset contains 96 numeric predictors to describe the
response variable. These predictors represent a wide range of financial
measures but can broadly be grouped into 5 categories:

-   Profitability

    -   Profitability measures describe a company's earnings relative to
        assets or equity and are variables like ROA A/B/C or Net Income
        to Total Assets.

-   Liquidity

    -   Liquidity measures the company's ability to turn assets into
        cash to meet short term financial obligations and are variables
        like Current Ratio or Workign Capital to Total Assets.

-   Leverage

    -   Leverage measures describe the company's debt and captial
        structure and are variables like Debt Ratio and Total Debt Total
        Net Worth

-   Turnover

    -   Turnover measures describe a company's efficiency of asset use
        and include variables like Cash Turnover Rate and Inventory
        Turnover Rate times.

-   Cash Flow

    -   Cash flow measures reflect the generation and allocatiion of
        cach within a company and include variables liek operting Cac
        Flow to Total Assets and Cash Current Liability.

::: {style="text-align: center;"}
### Missing Data
:::

```{r}
colSums(is.na(train))
```

One positive about the dataset is that there are no missing values. A
formal check showed that no missing values are present across any
predictors or the response variable. Therefore, no deletion or
augmentation of the data is needed.

::: {style="text-align: center;"}
### Collinearity Issues
:::

```{r}
library(caret)
library(dplyr)
num_train <- train[, sapply(train, is.numeric)]

nzv_idx <- nearZeroVar(num_train)
clean_train <- num_train[, -nzv_idx]

cor_mat <- cor(clean_train, use = "pairwise.complete.obs")

high_corr_pairs <- sum(
  abs(cor_mat[upper.tri(cor_mat)]) > 0.8,
  na.rm = TRUE
)

high_corr_pairs

top10_high_corr <- as.data.frame(as.table(cor_mat)) %>%
  filter(Var1 != Var2, abs(Freq) > 0.8) %>%
  mutate(abs_corr = abs(Freq)) %>%
  arrange(desc(abs_corr)) %>%
  slice(1:10)

top10_high_corr
```

There is significant collinearity among the predictors. Given the
financial nature of the dataset, this is expected. Many financial
figures and calculated metrics rely on core figures across the three
financial statements. The expectation is confirmed through correlation
analysis.

After accounting for undefined correlation caused by zero-variance
predictors, a pairwise correlation analysis shows that 43 distinct pairs
of predictors exhibit absolute correlation values greater than 0.8. Some
examples include Net Worth to Assets vs. Debt Ratio, and Operating Gross
Margin vs. Gross profit to Sales.

::: {style="text-align: center;"}
### Potential Important Predictors
:::

```{r}
cor_matrix <- cor(train, use = "complete.obs")
cor_response <- cor_matrix["Bankrupt.",]  # named vector

# Table of top correlated vars to response
data_frame(
  Variable = names(cor_response),
  Correlation = cor_response
) %>%
  filter(abs(Correlation) > 0.2) %>%
  arrange(desc(abs(Correlation)))

# Two sample T test
pvals <- sapply(train[ , -which(names(train) %in% c("X","Bankrupt."))], 
                function(x) {
                  tryCatch(t.test(x ~ train$Bankrupt.)$p.value, 
                           error = function(e) NA)
                })

sort(pvals)[1:15]
```

To identify some early signs of variables important in predicting
bankruptcy, we did both correlation analysis with the response variable
and two sample t-tests. Some notable variables narrowed down from both
results are as such:

Profitability measurement:

-   Net Income to Total Assets

-   ROA A/B/C before interest and after tax (these variables have high,
    near 1.0 correlation so either one of these variables could be
    selected)

Financial Leverage measurement:

-   Debt ratio

Liquidity measurement:

-   Working Capital to Total Assets

<!-- -->

-   Cash Total Assets

These early important variables include measures of profitability,
leverage, and liquidity that are early signals of bankruptcy risk and
may be important predictors later on.

::: {style="text-align: center;"}
### Key Figures and Numerical Summaries
:::

```{r}
summary(train$Bankrupt.)
prop.table(table(train$Bankrupt.))
```

The response variable Bankruptcy is a binary variable that takes values
of 0 or 1 to indicate if a company is or is not bankrupt. The numerical
summary of the response variable in the training dataset shows that
97.07% of the dataset is 0s while only 2.93% of the dataset is 1s. This
indicates that the dataset is very sparse and impalanced. This piece of
information is important to note when thinking about potential modeling
methods.

::: {style="text-align: center;"}
### Potential Models
:::

Overall, several numerical summaries capture the most important aspects
of the data. Severe class imbalance resulted in only 3% of the firms in
the training set being bankrupt, making it a rare event. There is
separation in financial performance, and profitability, leverage, and
earnings-related measures show strong differences between classes.
Extreme outliers contain implausible values and magnitudes larger than
expected. High redundancy exists, with predictors being nearly identical
or strongly correlated.

To overcome these issues based on the EDA, there should be
dimensionality reduction to address redundancy and the large number of
predictors. Examination of regularization or tree-based methods should
be prioritized over unregularized linear models. Finally, class
imbalance must be explicitly handled, and the selection of the best
model should emphasize sensitivity and recall rather than accuracy
alone.

::: {style="text-align: center;"}
## Model Analysis
:::

::: {style="text-align: center;"}
### Investigated Models, Description, and Assumptions
:::

Prior to our exploratory data analysis, we planned to train and test our
data using three methods: KNN, PLS, and Classification/Regression Trees.
However, after analyzing the structure of our data and evaluating
initial model performance, several issues became clear:

-   Extreme Class Imbalance: Bankruptcy cases make up only a small
    fraction of the dataset, causing these methods ot default to
    predicting the majority class.

-   High-Dimensional Numeric Ratios: The dataset contains many
    correlated financial ratios, which introduces noise and instability,
    especially harmful for KNN and PLS.

-   Nonlinear Relationships: The relationship between financial
    predictors and bankruptcy risk is highly nonlinear, limiting the
    effectiveness of simple tree models and linear projections.

-   Overlapping Distributions: Bankrupt and non-bankrupt have very
    similar predictor values, making it difficult for distance-based or
    linear-separation methods to distinguish between classes.

Given these challenges, we shifted our focus to models that are more
robust to imbalance, multicollinearity, and complex structure.
Specifically, we explored Ridge Regression, LASSO, Bagging, and Random
Forest, which provided significantly stronger performance and more
stable predictions. We examine these models and their results in the
following sections.

::: {style="text-align: center;"}
#### Ridge Regression
:::

```{r, include = FALSE}
# Setup Ridge Data
train_ridge <- train
test_ridge <- test

y_train_ridge <- as.numeric(as.character(train_ridge$Bankrupt.))
y_test_ridge <- as.numeric(as.character(test_ridge$Bankrupt.))

# Cleanup Columns
predictor_cols_ridge <- setdiff(colnames(train_ridge), c("Bankrupt.", "Unnamed..0"))

X_train_ridge <- as.matrix(train_ridge[, predictor_cols_ridge])
X_test_ridge <- as.matrix(test_ridge[, predictor_cols_ridge])

# Balanced Class Weights
class_counts <- table(train_ridge$Bankrupt.)
minority_weight <- class_counts["0"] / class_counts["1"]

ridge_weights <- ifelse(train_ridge$Bankrupt. == "1", minority_weight, 1)

# Fit Ridge Logistic Regression
ridge_cv <- cv.glmnet(
  x = X_train_ridge,
  y = y_train_ridge,
  alpha = 0,  
  family = "binomial",
  standardize = TRUE,
  weights = ridge_weights,
  type.measure = "deviance"
)

best_lambda_ridge <- ridge_cv$lambda.min
best_lambda_ridge

# Fit Final Ridge Model
ridge_fit <- glmnet(
  x = X_train_ridge,
  y = y_train_ridge,
  alpha = 0,
  family = "binomial",
  lambda = best_lambda_ridge,
  standardize = TRUE,
  weights = ridge_weights
)

# Predict on Test Data
ridge_prob <- predict(ridge_fit, newx = X_test_ridge, type = "response")

ridge_pred <- ifelse(ridge_prob > 0.5, "1", "0")
ridge_pred <- factor(ridge_pred, levels = c("0", "1"))

# Evaluate Ridge Performance
y_test_factor <- factor(test_ridge$Bankrupt., levels = c("0", "1"))
conf_mat_ridge <- confusionMatrix(ridge_pred, y_test_factor)
conf_mat_ridge

accuracy_ridge <- conf_mat_ridge$overall["Accuracy"]
sensitivity_ridge <- conf_mat_ridge$byClass["Sensitivity"]
specificity_ridge <- conf_mat_ridge$byClass["Specificity"]

accuracy_ridge
sensitivity_ridge
specificity_ridge
```

The first model we explored was Ridge Regression, which is well-suited
for datasets with highly correlated predictors, many numerical
variables, substantial noise, and class imbalance—all characteristics
that appeared in our exploratory analysis. Because Ridge is designed to
stabilize coefficient estimates in these settings, it was a strong
candidate for our bankruptcy prediction task.

Definition and Formula: Ridge Regression estimates the coefficients by
minimizing the usual sum of squared errors subject ot an L2 penalty that
shrinks the coefficients towards zero:

::: {style="text-align: center;"}
$\hat\beta_{Ridge} = argmin_{\beta} \sum_{i=1}^n (y_i - \beta_0 - \sum_{j=1}^p \beta_jx_{ij}) ^ 2$
subject to $\sum_{j=1}^p \beta_j^2 \le s$
:::

Here:

-   $\lambda \ge 0$ is the tuning parameter controlling shrinkage

-   Larger $\lambda$ forces coefficients toward 0 (reducing variance)

-   Ridge is useful when predictors are highlight correlated, numerous,
    or the model risks overfitting

Assumptions: Ridge Regression inherits the assumptions of logistic
regression, but relaxes some of them through its L2 penalty

1.  Linearity of log-odds: The model assumes a linear relationship
    between predictors and the log-odds of bankruptcy

    a.  Our check: Probability plots show some nonlinear patterns,
        meaning this assumption is not fully met

    b.  Why acceptable: Ridge is robust to mild misspecification because
        shrinkage stabilizes coefficients

2.  No perfect multicollinearity: Predictors should ot be perfectly
    correlated

    a.  Our check: Many financial ratios are highly correlated

    b.  Why acceptable: This is exactly what Ridge is designed for; L2
        regularization handles multicollinearity by shrinking correlated
        coefficients together

3.  Independent observations: Each row must represent an independent
    company

    a.  Our check: Firms appear only once, so this assumption is
        satisfied

4.  Reasonable class balance: Logistic regression is sensitive to
    imbalance

    a.  Our check: The data is extremely imbalanced

    b.  How addressed: We used class weights in the Ridge model to
        compensate

Assumption Summary: Although the dataset violates the linearity and
balance assumptions, Ridge Regression remains appropriate because the L2
penalty directly addresses multicollinearity, reduces variance, and
performs well in noisy, high-dimensional settings, as supported by its
strong AUC and sensitivity in our results.

After training the model and evaluating it on the test set, we obtained
the following confusion matrix:

```{r, echo = FALSE}
cm <- conf_mat_ridge$table
rownames(cm) <- c("Predicted Not Bankrupt", "Predicted Bankrupt")
colnames(cm) <- c("Actual Not Bankrupt", "Actual Bankrupt")
kable(cm, caption = "Ridge Regression Confusion Matrix")
```

From this output, we calculated several performance metrics:

-   Accuracy = 0.8583 - The model correctly classified about 86% of all
    firms in the test set.

-   Sensitivity = 0.8625 - The model correctly identified approximately
    86% of actual bankrupt companies, which is important given our goal
    of catching high-risk firms.

-   Specificity = 0.8581 - About 86% of non-bankrupt firms were
    correctly labeled, indicating strong performance on the majority
    class.

-   AUC = 0.9318 - This value shows that Ridge Regression is excellent
    at separating bankrupt vs. non-bankrupt firms across all
    classification thresholds, reflecting strong overall discriminative
    ability.

Overall, Ridge Regression performs well on this dataset and effectively
balances sensitivity and specificity, making it a strong foundation for
further analysis.

::: {style="text-align: center;"}
#### LASSO
:::

```{r, include = FALSE}
# Setup LASSO Data
train_lasso <- train
test_lasso  <- test

y_train_lasso <- as.numeric(as.character(train_lasso$Bankrupt.))
y_test_lasso  <- as.numeric(as.character(test_lasso$Bankrupt.))

predictor_cols_lasso <- setdiff(colnames(train_lasso), c("Bankrupt.", "Unnamed..0"))

X_train_lasso <- as.matrix(train_lasso[, predictor_cols_lasso])
X_test_lasso  <- as.matrix(test_lasso[, predictor_cols_lasso])

# Balanced Class Weights
class_counts <- table(train_lasso$Bankrupt.)
minority_weight <- class_counts["0"] / class_counts["1"]

lasso_weights <- ifelse(train_lasso$Bankrupt. == "1", minority_weight, 1)

# Fit LASSO Logistic Regression
lasso_cv <- cv.glmnet(
  x = X_train_lasso,
  y = y_train_lasso,
  alpha = 1,
  family = "binomial",
  standardize = TRUE,
  weights = lasso_weights, 
  type.measure = "deviance"
)

best_lambda_lasso <- lasso_cv$lambda.min
best_lambda_lasso

# Fit Final Lasso Model
lasso_fit <- glmnet(
  x = X_train_lasso,
  y = y_train_lasso,
  alpha = 1,
  family = "binomial",
  lambda = best_lambda_lasso,
  standardize = TRUE,
  weights = lasso_weights
)

# Predict on Test Data
lasso_prob <- predict(lasso_fit, newx = X_test_lasso, type = "response")

lasso_pred <- ifelse(lasso_prob > 0.5, "1", "0")
lasso_pred <- factor(lasso_pred, levels = c("0", "1"))

# Evaluate LASSO Performance
y_test_factor_lasso <- factor(test_lasso$Bankrupt., levels = c("0", "1"))

conf_mat_lasso <- confusionMatrix(lasso_pred, y_test_factor_lasso)
conf_mat_lasso

accuracy_lasso    <- conf_mat_lasso$overall["Accuracy"]
sensitivity_lasso <- conf_mat_lasso$byClass["Sensitivity"]
specificity_lasso <- conf_mat_lasso$byClass["Specificity"]

accuracy_lasso
sensitivity_lasso
specificity_lasso
```

The next model we evaluated was LASSO Regression, which is particularly
useful for datasets with many predictors and potential redundancy.
Unlike Ridge, which shrinks all coefficients, LASSO can set some
coefficients to zero, effectively performing feature selection. Given
the high dimensionality of the financial ratios in our dataset, LASSO
was a strong candidate because it helps identify the most important
predictors while reducing noise.

Definition and Formula: LASSO (Least Absolute Shrinkage and Selection
Operator) estimates the coefficients by minimizing the sum of squared
errors subject to an L1 penalty, which both shrinks coefficients and can
set some exactly to zero (performing variable selection)

::: {style="text-align: center;"}
$\hat\beta_{LASSO} = argmin_{\beta} \sum_{i=1}^n (y_i - \beta_0 - \sum_{j=1}^p \beta_jx_{ij}) ^ 2$
subject to $\sum_{j=1}^p |\beta_j| \le s$
:::

Here:

-   $\lambda \ge 0$ controls the amount of shrinkage

-   The L1 penalty forces some coefficients exactly to zero

-   LASSO is especially useful when feature selection is desired or when
    many predictors are irrelevant

Assumptions: LASSO relies on the same core assumptions as logistic
regression, but its L1 penalty adds sparsity and variable selection.

1.  Linearity of the log-odds: The model assumes a linear relationship
    between predictors and the log-odds of bankruptcy

    a.  Our check: Probability plots showed some nonlinear patterns, so
        this assumption is not perfectly met

    b.  Why acceptable: LASSO's regularization helps reduce overfitting
        even when the relationship is not perfectly linear

2.  No perfect multicollinearity: Predictors should not be perfectly
    correlated

    a.  Our check: Financial ratios are strongly correlated

    b.  Issue: LASSO tends to pick one variable from a correlated group
        and discard the rest, which can create instability.

    c.  Why acceptable: This behavior is actually useful for variable
        selection, and the model still performed well in testing.

3.  Independent observations: Each observation should be independent

    a.  Our check: Each company appears once, so this assumption is
        satisfied

4.  Class balance: Logistic-based models can struggle with extreme
    imbalance

    a.  Our check: The dataset is highly imbalanced

    b.  How addressed: Class weights were used to reduce bias toward the
        majority class

Assumption Summary: LASSO does not perfectly satisfy linearity or
balance assumptions, and its variable selection can be unstable under
multicollinearity. However, it still performed strongly and produced a
sparse, interpretable model, making it a useful complementary approach,
even if not the final chosen model.

After fitting the LASSO model on the training set and evaluating it on
the test set, we obtained the following confusion matrix:

```{r, echo = FALSE}
cm <- conf_mat_lasso$table
rownames(cm) <- c("Predicted Not Bankrupt", "Predicted Bankrupt")
colnames(cm) <- c("Actual Not Bankrupt", "Actual Bankrupt")
kable(cm, caption = "LASSO Confusion Matrix")
```

From this, we computed the key performance metrics:

-   Accuracy = 0.8612 - The model correctly classified about 86% of the
    companies in the test set.

-   Sensitivity = 0.8500 - LASSO correctly identified about 85% of the
    actual bankrupt companies.

-   Specificity = 0.8616 - About 86% of the non-bankrupt companies were
    correctly classified.

-   AUC = 0.9301 - This indicates excellent class separation, nearly as
    strong as Ridge, demonstrating that LASSO is also highly effective
    at distinguishing between bankrupt and non-bankrupt firms across all
    thresholds.

Overall, LASSO performed very similarly to Ridge, with strong accuracy
and discrimination ability. Its slightly lower sensitivity compared to
Ridge suggests it identifies slightly fewer bankrupt firms, but its
interpretability advantage (automatic variable selection) makes it
especially useful for understanding which financial ratios matter most.

::: {style="text-align: center;"}
#### Bagging
:::

```{r, include = FALSE}
# Setup Bagging Data
train_bag <- train
test_bag  <- test

train_bag$Bankrupt. <- factor(train_bag$Bankrupt., levels = c("0","1"))
test_bag$Bankrupt.  <- factor(test_bag$Bankrupt., levels = c("0","1"))

predictor_cols <- setdiff(colnames(train_bag), c("Bankrupt.", "Unnamed..0"))

X_train <- train_bag[, predictor_cols]
y_train <- train_bag$Bankrupt.

X_test  <- test_bag[, predictor_cols]
y_test  <- test_bag$Bankrupt.

# Balanced Class Weights
class_counts <- table(y_train)
minority_size <- min(class_counts) 
sampsize_vec <- rep(minority_size, length(class_counts))

# Train Model
bag_model <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  mtry = ncol(X_train), 
  sampsize = sampsize_vec,
  importance = TRUE
)

# Predict on Test Data
bag_prob <- predict(bag_model, X_test, type = "prob")[,2]

bag_pred <- factor(ifelse(bag_prob > 0.5, "1", "0"), levels = c("0","1"))

# Evaluate Model
conf_mat_bag <- confusionMatrix(
  bag_pred,
  y_test,
  positive = "1"
)

conf_mat_bag

accuracy_bag     <- conf_mat_bag$overall["Accuracy"]
sensitivity_bag  <- conf_mat_bag$byClass["Sensitivity"]
specificity_bag  <- conf_mat_bag$byClass["Specificity"]

accuracy_bag
sensitivity_bag
specificity_bag
```

Bagging is an ensemble method that reduces variance by training many
decision trees on bootstrapped samples of the data and averaging their
predictions. Because individual trees are highly sensitive to noise and
instability, bagging helps stabilise predictions, making it a strong
candidate for noisy, high-dimensional, and imbalanced datasets like
ours.

Definition and Formula: Bagging is an ensemble learning method that
reduces variance by training multiple models on different bootstrap
samples of the data and averaging their predictions:

::: {style="text-align: center;"}
$\hat f_{bag}(x) = \frac{1}{B}\sum_{b=1}^B \hat f_b(x)$
:::

Where:

-   B is the number of bootstrap samples

-   Each $\hat f_b(x)$ is the model trained on the b-th bootstrap sample

-   The final prediction is the average of all models for regression or
    a majority vote for classification

Assumptions: Bagging relies on far fewer assumptions than statistical
models like logistic regression. Since it uses machine-learning decision
trees trained on bootstrapped samples, it is flexible, nonparametric,
and robust to complex nonlinear relationships.

1.  No linearity assumption: Decision trees do not require linear
    relationships between predictors and the outcome

    a.  Our check: Our exploratory plots showed nonlinear trends in
        several financial ratios, making Bagging appropriate

2.  Handles multicollinearity: Tree-based models are not affected by
    strongly correlated predictors

    a.  Our check: Many ratios in our dataset were highly correlated,
        and Bagging naturally manages this without instability

3.  Handles outliers and skewed variables: Decision trees partition data
    based on thresholds, so extreme values do not distort the model

    a.  Our check: Several financial ratios were heavily skewed; Bagging
        remains reliable under these conditions

4.  No distributional assumptions: Bagging does not assume normality or
    equal variances

    a.  Our check: These assumptions were violated in the data, but
        Bagging does not require them

5.  Requires reasonably independent observations: Observations should
    not be duplicated or repeated entities

    a.  Our check: Each fire is observed once, so this assumption is met

Assumption Summary: Although Bagging does not directly correct the
extreme class imbalance, we addressed this using balanced sampling

When we evaluated the Bagging model on the test set, the confusion
matrix was:

```{r, echo = FALSE}
cm <- conf_mat_bag$table
rownames(cm) <- c("Predicted Not Bankrupt", "Predicted Bankrupt")
colnames(cm) <- c("Actual Not Bankrupt", "Actual Bankrupt")
kable(cm, caption = "Bagging Confusion Matrix")
```

From this output, we obtained the following metrics:

-   Accuracy = 0.8876 - The model correctly classified about 89% of all
    companies in the test set.

-   Sensitivity = 0.7625 - Bagging identified roughly 76% of bankrupt
    companies, which is strong given the severe class imbalance.

-   Specificity = 0.8927 - Nearly 89% of non-bankrupt firms were
    correctly labelled.

-   AUC = 0.9281 - This high AUC indicates that Bagging separates
    bankrupt vs. non-bankrupt companies well across all probability
    thresholds.

Overall, Bagging performed very well, achieving a strong balance between
sensitivity and specificity, making it one of our top-performing models.

::: {style="text-align: center;"}
#### Random Forest
:::

```{r, include = FALSE}
# Setup Random Forest Data
train_rf <- train
test_rf  <- test

train_rf$Bankrupt. <- factor(train_rf$Bankrupt., levels = c("0","1"))
test_rf$Bankrupt.  <- factor(test_rf$Bankrupt.,  levels = c("0","1"))

predictor_cols <- setdiff(colnames(train_rf), c("Bankrupt.", "Unnamed..0"))

X_train <- train_rf[, predictor_cols]
y_train <- train_rf$Bankrupt.

X_test  <- test_rf[, predictor_cols]
y_test  <- test_rf$Bankrupt.

# Balanced Sampling Setup
class_counts <- table(y_train)
minority_size <- min(class_counts)

sampsize_vec <- rep(minority_size, length(class_counts))

# Fit Random Forest
rf_model <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  mtry = floor(sqrt(ncol(X_train))),
  sampsize = sampsize_vec, 
  importance = TRUE
)

# Predict on Test Set
rf_prob <- predict(rf_model, X_test, type = "prob")[,2]

rf_pred <- factor(ifelse(rf_prob > 0.5, "1", "0"), levels = c("0","1"))

# Confusion Matrix and Metrics
conf_mat_rf <- confusionMatrix(
  rf_pred,
  y_test,
  positive = "1"
)

conf_mat_rf

accuracy_rf     <- conf_mat_rf$overall["Accuracy"]
sensitivity_rf  <- conf_mat_rf$byClass["Sensitivity"]
specificity_rf  <- conf_mat_rf$byClass["Specificity"]

accuracy_rf
sensitivity_rf
specificity_rf
```

Random Forest builds on the idea of Bagging by adding an additional
layer of randomness. Each tree in the ensemble is trained on a
bootstrapped sample and only a random subset of predictors at each
split. This helps decorrelate the trees, reduces overfitting, and often
improves performance, especially in datasets with many correlated
financial ratios, like ours.

Definition and Formula: Random Forest is an ensemble learning method
that builds multiple decision trees using both bootstrap samples of the
data and random subsets of predictors at each split, reducing variance
and limiting overfitting:

::: {style="text-align: center;"}
$\hat f_{RF}(x) = \frac{1}{B}\sum_{b=1}^B T_b(x)$
:::

Where:

-   B is the number of trees

-   $T_b(x)$ is the prediction from the b-th decision tree

-   Each tree is trained on a bootstrap sample and, at each split,
    chooses from a random subset of predictors rather than all
    predictors

Assumptions: Random Forest is a nonparametric ensemble method with far
fewer assumptions than linear models, making it well-suited for noisy,
high-dimensional financial data.

1.  Independence of observations: Each row should represent an
    independent data point

    a.  Our Check: Each firm appears once in the dataset, so this
        assumption is satisfied

2.  Sufficient variability and sample size: Random Forest requires
    enough data for bootstrapping to create diverse trees

    a.  Our Check: With thousands of firms and many predictors, the
        dataset provides ample variation

    b.  Why Acceptable: The model produced stable out-of-bag error
        estimates, indicating enough data for reliable tree construction

3.  No assumptions of linearity or normality: Random Forest does not
    require linear relationships, normally distributed predictors, or
    homoscedasticity

    a.  Our Check: Financial ratios showed strong nonlinear
        relationships and skewed distributions

    b.  Why Acceptable: These patterns are expected in bankruptcy data,
        and Random Forest naturally captures nonlinear interactions

4.  Handles multicollinearity automatically: Tree-based models are not
    affected by correlated predictors

    a.  Our Check: Many features were heavily correlated

    b.  Why Acceptable: Random feature selection prevents trees from
        repeatedly splitting on the same correlated variables, reducing
        overfitting

Assumption Summary: Random Forest satisfies its core assumptions and is
well-matched to this dataset's structure: nonlinear relationships,
correlated financial ratios, and high dimensionality.

After training the Random Forest model on our dataset, the confusion
matrix was:

```{r, echo = FALSE}
cm <- conf_mat_rf$table
rownames(cm) <- c("Predicted Not Bankrupt", "Predicted Bankrupt")
colnames(cm) <- c("Actual Not Bankrupt", "Actual Bankrupt")
kable(cm, caption = "Random Forest Confusion Matrix")
```

From this model output, we computed the following evaluation metrics:

-   Accuracy = 0.8895 - Random Forest correctly classified about 89% of
    firms in the test set, tying Bagging for the highest overall
    accuracy.

-   Sensitivity = 0.7625 - The model detected about 76% of actual
    bankrupt companies, performing slightly below Bagging in the recall
    of minority cases.

-   Specificity = 0.8947 - Nearly 89% of non-bankrupt firms were
    correctly identified, the strongest specificity among all models.

-   AUC = 0.9305 - A high AUC score indicates strong ranking ability and
    good separation between bankrupt and non-bankrupt companies.

Overall, Random Forest performed extremely well, offering one of the
best combinations of accuracy, specificity, and AUC among the models
tested. It's a slight trade-off in sensitivity compared to Bagging,
which is common, but it remained one of the strongest-performing
classifiers on this dataset.

::: {style="text-align: center;"}
#### Model Comparison
:::

```{r, warning = FALSE, include = FALSE}
evaluate_model <- function(probabilities, true_labels, threshold = 0.5) {
  
  preds <- factor(ifelse(probabilities > threshold, "1", "0"), levels = c("0","1"))
  
  cm <- confusionMatrix(preds, true_labels, positive = "1")
  
  roc_obj <- roc(true_labels, probabilities)
  auc_val <- as.numeric(auc(roc_obj)) 
  
  tibble(
    Accuracy = as.numeric(cm$overall["Accuracy"]),
    Sensitivity = as.numeric(cm$byClass["Sensitivity"]),
    Specificity = as.numeric(cm$byClass["Specificity"]),
    Balanced_Accuracy = as.numeric((cm$byClass["Sensitivity"] + cm$byClass["Specificity"]) / 2),
    AUC = auc_val
  )
}

ridge_results <- evaluate_model(ridge_prob, y_test_factor)
lasso_results <- evaluate_model(lasso_prob, y_test_factor)
bagging_results <- evaluate_model(bag_prob, y_test_factor)
rf_results <- evaluate_model(rf_prob, y_test_factor)

comparison_table <- bind_rows(
  Ridge = ridge_results,
  LASSO = lasso_results,
  Bagging = bagging_results,
  RandomForest = rf_results,
  .id = "Model"
)
```

After running all four models, we compared their performance using the
key metrics discussed earlier:

```{r, warning = FALSE, echo = FALSE}
kable(comparison_table, digits = 4, caption = "Model Performance Comparison")
```

Across the models, performance was similar, but each model showed
strengths in different areas. Random Forest achieved the highest overall
accuracy and specificity, indicating it excelled at correctly
identifying non-bankrupt companies. Bagging performed similarly well in
these categories, followed by LASSO and Ridge. Since Random Forest
performed well on both accuracy and specificity, it also achieved one of
the second-highest AUC values among the models.

However, when focusing on sensitivity (the model’s ability to correctly
identify companies that actually went bankrupt), Ridge Regression
performed the best. Ridge correctly flagged the largest proportion of
bankrupt firms, making it the strongest model for detecting the rare but
critical positive class. Ridge also achieved the highest overall AUC,
indicating that it best separates bankrupt from non-bankrupt companies
across all probability thresholds.

While the four models are comparable overall, the choice depends on the
relative cost of different errors. In a scenario where failing to
identify a company at risk of bankruptcy is far more costly than
predicting a false alarm, sensitivity becomes the most important metric.
Because Ridge Regression not only maximizes sensitivity but also yields
the highest AUC, it represents the most reliable and balanced model for
our prediction task. In addition, the diagnostic plots, particularly the
calibration curve and predicted probability distributions, show no major
violations of Ridge Regression assumptions, further supporting its
appropriateness for this dataset.

Therefore, the final model selected was Ridge Regression.

::: {style="text-align: center;"}
#### Challenges Faced
:::

Throughout the modeling process, we encountered several challenges with
the bankruptcy dataset's structure. The data was extremely imbalanced,
with very few bankrupt firms, which made sensitivity a key concern and
caused many baseline models to misclassify high-risk companies. In
addition, the financial predictors exhibited strong multicollinearity
and nonlinear relationships, violating the assumptions of traditional
logistic models. Ridge Regression helped address these issues because
its L2 penalty stabilizes coefficients in the presence of correlated
variables, and its performance diagnostics suggested no major deviations
from logistic model assumptions after regularization.

We also explored several alternative models, each of which revealed
specific limitations under these conditions. KNN struggled heavily with
the curse of dimensionality and the imbalance problem, leading to poor
minority-class detection. PLS regression handled multicollinearity but
could not capture important nonlinear relationships in the data.
Classification trees were highly unstable and prone to overfitting,
especially given the noise and imbalance in the predictors. By
evaluating these challenges and applying class balancing,
regularization, and diagnostic validation, we determined that Ridge
Regression offered the most reliable combination of accuracy,
interpretability, and robustness. Ultimately, this made it the strongest
model for predicting bankruptcy risk in this dataset.

::: {style="text-align: center;"}
### Model Validation
:::

To confirm our decision of the Ridge Regression model, we decided to
explore the details of the model to truly understand how it works.

::: {style="text-align: center;"}
```{r, message = FALSE, warning = FALSE, echo = FALSE}
ridge_roc <- roc(y_test_ridge, as.numeric(ridge_prob))
#auc(ridge_roc)
plot(ridge_roc, main = "Ridge Regression ROC Curve")
```
:::

The Ridge Regression model achieved an AUC of .9318, indicating
excellent ability to distinguish between bankrupt and non-bankrupt
firms. An AUC this high means the model correctly ranks risky companies
above healthy ones the vast majority of the time, even across different
probability thresholds. This strong discriminative performance is
especially noteworthy given the dataset's extreme class imbalance,
demonstrating that Ridge Regression is highly effective at identifying
high-risk firms despite the data's challenging structure.

::: {style="text-align: center;"}
```{r, warning = FALSE, echo = FALSE}
ridge_coefs <- coef(ridge_fit)

coef_df <- data.frame(
  Variable = rownames(ridge_coefs),
  Coefficient = as.numeric(ridge_coefs)
) %>%
filter(Variable != "(Intercept)") %>%
mutate(Abs_Coefficient = abs(Coefficient)) %>%
arrange(desc(Abs_Coefficient))

coef_df_top <- coef_df %>% slice_max(Abs_Coefficient, n = 20)

ggplot(coef_df_top, aes(x = reorder(Variable, Coefficient), y = Coefficient)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 20 Ridge Regression Coefficients (by Magnitude)",
    x = "Predictor",
    y = "Coefficient"
  ) +
  theme_minimal()
```
:::

The top 20 Ridge Regression coefficients highlight which financial
ratios have the strongest influence on bankruptcy after shrinkage is
applied. Variables with large positive coefficients, such as Debt Ratio
and Current Liability to Assets, increase the predicted probability of
bankruptcy, indicating they are strong risk indicators. In contrast,
variables with large negative coefficients, such as Persistent EPS in
the Last Four Seasons and Contingent Liabilities to Net Worth, reduce
predicted bankruptcy risk. Since Ridge shrinks correlated predictors
together rather than eliminating them, this ranking reflects the overall
combined importance of financial ratios rather than selecting just a few
individual features.

::: {style="text-align: center;"}
```{r, warning = FALSE, echo = FALSE}
ridge_prob_df <- data.frame(
  Predicted_Prob = as.numeric(ridge_prob),
  Actual = factor(y_test_ridge, levels = c("0", "1"))
)

ggplot(ridge_prob_df, aes(x = Predicted_Prob, fill = Actual)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Ridge Regression Predicted Probability Distribution",
    x = "Predicted Bankruptcy Probability",
    fill = "Actual Class"
  ) +
  theme_minimal()
```
:::

This probability distribution plot shows how well the Ridge Regression
model separates bankrupt and non-bankrupt companies based on predicted
risk scores. Non-bankrupt firms (Actual Class 0) are mostly concentrated
near very low predicted probabilities, while bankrupt firms (Actual
Class 1) cluster toward the higher end of the probability scale.
Although there is some overlap in the middle range, as expected given
the noisy, imbalanced data, the two distributions remain well-separated.
This indicates that the model assigns meaningfully different risk levels
to the two groups and effectively ranks firms by bankruptcy likelihood.

::: {style="text-align: center;"}
```{r, warning = FALSE, echo = FALSE}
cal_data <- data.frame(
  prob = as.numeric(ridge_prob),
  actual = as.numeric(as.character(y_test_ridge))
)

cal_data$bin <- cut(cal_data$prob, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)

cal_summary <- cal_data %>%
  group_by(bin) %>%
  summarize(
  mean_pred = mean(prob),
  actual_rate = mean(actual)
)

ggplot(cal_summary, aes(x = mean_pred, y = actual_rate)) +
  geom_point(size = 3, color = "steelblue") +
  geom_line(color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Ridge Regression Calibration Curve",
    x = "Average Predicted Probability",
    y = "Observed Bankruptcy Rate"
  ) +
  theme_minimal()

```
:::

The calibration curve evaluates how well Ridge Regression's predicted
probabilities match actual bankruptcy rates. Ideally, points would lie
on the diagonal dashed line, meaning predicted risks equal observed
outcomes. In our plot, the model is slightly underconfident at lower
probabilities (predicting small but nonzero risk even when almost no
firms go bankrupt) and becomes more accurate as predicted probabilities
increase, with strong alignment in the highest-risk bins. This pattern
indicates that while Ridge tends to "play it safe" at the low end, it
reliably indicates genuinely high-risk firms, making its probability
estimates useful for ranking and decision-making.

::: {style="text-align: center;"}
```{r, warning = FALSE, echo = FALSE}
lift_data <- gains(
  actual = as.numeric(as.character(y_test_ridge)),
  predicted = as.numeric(ridge_prob),
  groups = 10
)

lift_df <- data.frame(
  depth = lift_data$depth,
  lift = lift_data$lift
)

ggplot(lift_df, aes(x = depth, y = lift)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Ridge Regression Lift Chart",
    x = "Percentile of Sample",
    y = "Lift"
  ) +
  theme_minimal()
```
:::

The lift chart shows how effectively Ridge Regression identifies the
highest-risk companies. The very first segment of the ranked predictions
achieves extremely high lift, meaning the model concentrates a
disproportionately large share of actual bankruptcies among the firms it
ranks as most likely to fail. Lift drops sharply after this top group,
and later segments approach the performance of random selection. This
pattern indicates that Ridge is highly effective at prioritizing the
riskiest companies, even if its probability calibration is not perfect.

```{r, warning = FALSE, echo = FALSE}
top_risk <- data.frame(
  Index = 1:length(ridge_prob),
  Predicted_Prob = as.numeric(ridge_prob),
  Actual = y_test_ridge
) %>%
arrange(desc(Predicted_Prob)) %>%
head(10)

kable(top_risk, digits = 4, caption = "Top 10 Highest-Risk Companies")
```

The table lists the 10 companies with the highest predicted probability
of bankruptcy according to Ridge Regression. All ten firms have
extremely high predicted probabilities, and most importantly, every
company in this group is actually labeled bankrupt in the test set. This
demonstrates that the model is highly effective at identifying the most
at-risk firms. When Ridge assigns a near-certain probability of
bankruptcy, it is consistently correct. This strong alignment between
predicted risk and true outcomes reinforces the model's value for
early-warning screening and prioritizing firms that require immediate
attention.

::: {style="text-align: center;"}
## Results
:::

To evaluate the final model performance, we tested the Ridge Regression
model on our test dataset. Ridge was selected because it demonstrated
strong discriminative performance, stability under multicollinearity,
and effective handling of the highly imbalanced class structure.

Across the test set, Ridge achieved an accuracy of 0.8583, a sensitivity
of 0.8625, and a specificity of 0.8581. Below are interpretations of
these values in the context of the problem:

-   85.83% of the time, the Ridge Regression model will correctly
    identify if a company will go bankrupt or not

-   86.25% of the time, the Ridge Regression model will correctly
    identify a company that will go bankrupt

-   85.81% of the time, the Ridge Regression model will correctly
    identify a company that will not go bankrupt

Importantly, Ridge produced an AUC of .9318, indicating excellent
ability to rank companies by bankruptcy risk. Below is the
interpretation of this value in the context of the problem:

-   The Ridge Regression model has a 93.18% chance of correctly ranking
    a randomly chosen bankrupt firm above a randomly chosen non-bankrupt
    firm

The model consistently assigns very low probabilities to healthy firms
and high probabilities to bankrupt firms, as shown in the validation
probability distribution plot.

Interpreting the final fitted model, the largest coefficient magnitudes
indicate which financial ratios contribute most strongly to bankruptcy
risk. Variables such as Debt Ratio, Current Liability to Assets, and No
Credit Interval increase the predicted bankruptcy probability, while
metrics such as Persistent EPS over the Last Four Seasons and Contingent
Liabilities to Net Worth reduce risk. Since Ridge shrinks correlated
predictors together, their rankings reflect overall risk signals rather
than reliance on a single feature.

Finally, we examined the model's highest-risk predictions. The top 10
firms with the highest predicted risk in the test set were all labeled
as bankrupt, demonstrating that Ridge is highly effective at identifying
the most vulnerable companies. This makes the model particularly useful
for early warning systems and for prioritizing firms for further
financial assessment.

Overall, Ridge Regression provides strong predictive accuracy, excellent
ranking performance, and interpretable financial drivers of bankruptcy,
making it a reliable model for this application.

::: {style="text-align: center;"}
## Conclusion
:::

In this analysis, we developed and evaluated several predictive models
for bankruptcy classification and ultimately selected Ridge Regression
as the final model due to its strong performance, stability under
multicollinearity, and reliable handling of highly imbalanced financial
data. Ridge achieved high accuracy and an AUC above .93, demonstrating
excellent ability to distinguish between bankrupt and non-bankrupt
firms, while its coefficient structure provided meaningful insight into
the financial ratios most associated with bankruptcy risk. Model
validation further showed that Ridge effectively ranks firms by risk and
reliably identifies the highest-risk companies. Overall, the Ridge
Regression model offers a strong balance of predictive accuracy,
interpretability, and robustness, making it well-suited for risk
assessment applications.
