---
title: 'Part 1: EDA'
author: "Angela Wei, Aditya Desai"
date: "2025-12-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(ggplot2)

training_df <- read_csv("../data/train.csv")

```

EDA Instructions:

First, introduce the dataset and perform a careful EDA of the data. Consider the following
questions to help motivate your EDA:
• Are there any problems with the data?
• What kind of variables do you have?
• Is there any missing data?
• Does there appear to be potentially problematic collinearity amongst the predictor variables?
Can you detect early signs of what variables are likely to be important in predicting the
response?
• What are the key figures or numerical summaries that describe the most important aspects
of the data?
• Does your EDA suggest to you what modeling approaches you should aim to try?
Consider these questions and any others that come to mind as you familiarize yourself with the
data.

# EDA

```{r}
head(training_df)
```

