---
title: 'Part 1: EDA'
author: "Angela Wei, Aditya Desai"
date: "2025-12-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(ggplot2)
library(dplyr)

training_df <- read_csv("../data/train.csv")

```

• Does there appear to be potentially problematic collinearity amongst the predictor variables?
Can you detect early signs of what variables are likely to be important in predicting the
response?
• What are the key figures or numerical summaries that describe the most important aspects
of the data?
• Does your EDA suggest to you what modeling approaches you should aim to try?
Consider these questions and any others that come to mind as you familiarize yourself with the
data.

# Introducing dataset

This dataset is about predicting whether a company will go bankrupt for not given explanatory variables. The data is collected from Taiwan Economic Journal from 1999-2009 so company bankruptcy is thus defined based on the Taiwan stock exchange. 

The response variable is `Bankrupt.`, a binary classification of bankruptcy status. There are 95 continuous explanatory variables. The predictor variables are some financial ratios, company performance metrics, and more. 
```{r}
# Preview of Dataset
head(training_df)
colnames(training_df)

# It seems like the `...1` column is a meaningless automativally generated index value, not seen in the kaggle data description, so I remove it.
head(training_df$...1)
training_df <- training_df %>%
  select(-...1)
```

# Data Quality Check
```{r}
# Check for duplicated values
sum(duplicated(training_df))

# Check for missing values
sum(is.na(training_df))

# Check that the response var. is the only categorical var.
unique(training_df$Bankrupt.)
unname(sapply(training_df, class))

# Check for negative values or ratios that are greater than 1
neg_values <- sapply(training_df,function(x) any(x < 0, na.rm = TRUE))
sum(neg_values)

ratio_candidates <- sapply(training_df, function(x) all(x >= 0 & x <= 1, na.rm = TRUE))
names(ratio_candidates[ratio_candidates == TRUE])
names(ratio_candidates[ratio_candidates == FALSE])
```

There is no duplicated rows or missing data in the dataset. The response variable is valued only 0 or 1 as it should be and all other 95 variables are numeric variables. None of the variables have negative values or values which wouldn't make sense in context of the variable measurement.




# 
