---
title: 'Part 1: EDA'
author: "Angela Wei, Aditya Desai"
date: "2025-12-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(ggplot2)
library(dplyr)
library(rlang)
library(tidyverse)


training_df <- read_csv("../data/train.csv")

```



# Introducing dataset

This dataset is about predicting whether a company will go bankrupt for not given explanatory variables. The data is collected from Taiwan Economic Journal from 1999-2009 so company bankruptcy is thus defined based on the Taiwan stock exchange. 

The response variable is `Bankrupt.`, a binary classification of bankruptcy status. There are 95 continuous explanatory variables. The predictor variables are some financial ratios, company performance metrics, and more. 
```{r}
# Preview of Dataset
head(training_df)
# It seems like the `...1` column is a meaningless automativally generated index value, not seen in the kaggle data description, so I removed it.
head(training_df$...1)
training_df <- training_df %>%
  select(-...1)
```

# Data Quality Check
```{r}
# Check for duplicated values
sum(duplicated(training_df))

# Check for missing values
sum(is.na(training_df))

# Check that the response var. is the only categorical var.
unique(training_df$Bankrupt.)
unname(sapply(training_df, class))

# Check for negative values or ratios that are greater than 1
neg_values <- sapply(training_df,function(x) any(x < 0, na.rm = TRUE))
sum(neg_values)

ratio_candidates <- sapply(training_df, function(x) all(x >= 0 & x <= 1, na.rm = TRUE))
names(ratio_candidates[ratio_candidates == TRUE])
names(ratio_candidates[ratio_candidates == FALSE])
```

There is no duplicated rows or missing data in the dataset. The response variable is valued only 0 or 1 as it should be and all other 95 variables are numeric variables. None of the variables have negative values but there are some variables that are rates or ratios that seem to have very high values.

# Variables EDA
```{r}
# Select all covariates & plot them
vars <- colnames(training_df)[-1]
for (var in vars) {
  p <- ggplot(training_df, aes_string(y = var)) +
    geom_boxplot(fill = "pink", outlier.color = "red") +
    coord_flip() +
    ggtitle(paste("Boxplot of", var))
  print(p)
}
```
```{r}
run_cor_test <- function(x, y) {
  test <- cor.test(x, y)
  tibble(
    correlation = unname(test$estimate),
    p_value = test$p.value
  )
}

# Run cor.test on all predictors
cor_test_results <- training_df %>%
  select(-Bankrupt.) %>%
  map_dfr(~ run_cor_test(.x, training_df$Bankrupt.)) %>%
  mutate(variable = names(training_df)[names(training_df) != "Bankrupt."]) %>%
  select(variable, correlation, p_value) %>%
  arrange(desc(abs(correlation)))

# View results
temp <- head(cor_test_results, 15)
temp$variable
temp
```

Tree based importance stuff

```{r}
library(randomForest)

# Make response a factor for classification
training_df$Bankrupt. <- as.factor(training_df$Bankrupt.)

# Fit a random forest
rf_model <- randomForest(
  Bankrupt. ~ ., 
  data = training_df,
  importance = TRUE
)

# Extract feature importance
importance_df <- as.data.frame(rf_model$importance)

# Sort by MeanDecreaseGini (or MeanDecreaseAccuracy)
importance_df$variable <- rownames(importance_df)
importance_df <- importance_df[order(-importance_df$MeanDecreaseGini), ]

head(importance_df, 10)
```


```{r}
run_cor_test <- function(x, y) {
  test <- biserial.cor(x, y)
  tibble(
    correlation = unname(test$estimate),
    p_value = test$p.value
  )
}

# Run cor.test on all predictors
cor_test_results <- training_df %>%
  select(-Bankrupt.) %>%
  map_dfr(~ run_cor_test(.x, training_df$Bankrupt.)) %>%
  mutate(variable = names(training_df)[names(training_df) != "Bankrupt."]) %>%
  select(variable, correlation, p_value) %>%
  arrange(desc(abs(correlation)))

# View results
temp <- head(cor_test_results, 15)
temp$variable
temp
```


# Predictor collinearity
```{r}
# Predictors with high correlation to each other
cor_matrix <- cor(training_df[, vars], use = "complete.obs")

# 0.8 threshold for high correlation
high_corr <- which(abs(cor_matrix) > 0.8 & abs(cor_matrix) < 1, arr.ind = TRUE)
high_corr <- high_corr[high_corr[,1] < high_corr[,2], ]

# Table the variable pairs
data.frame(
  Var1 = rownames(cor_matrix)[high_corr[,1]],
  Var2 = colnames(cor_matrix)[high_corr[,2]],
  Correlation = cor_matrix[high_corr]
) %>% 
  arrange(desc(Correlation))
```
There are a lot of highly correlated variables with Operatig Gross Margin having perfect correlation with Gross Profit to Sales. This indicates that a lot of these variables are redundant, essentially measuring the same concept. 

# Variable correlation to response

```{r}
# Compute correlation matrix & take out correlations with Bankrupt.
cor_matrix <- cor(training_df, use = "complete.obs")
cor_response <- cor_matrix["Bankrupt.",]  # named vector
cor_response

# Table of top correlated vars to response
data_frame(
  Variable = names(cor_response),
  Correlation = cor_response
) %>%
  filter(abs(Correlation) > 0.2) %>%
  arrange(desc(abs(Correlation)))
```
```{r}
pvals <- sapply(training_df[ , -which(names(training_df) %in% c("X","Bankrupt."))], 
                function(x) {
                  tryCatch(t.test(x ~ training_df$Bankrupt.)$p.value, 
                           error = function(e) NA)
                })

sort(pvals)[1:15]
```

# Ending thoughts
```{r}
ggplot(training_df, aes(x = factor(Bankrupt.), y = Debt.ratio.., fill = factor(Bankrupt.))) +
  geom_boxplot() +
  labs(x = "Bankrupt", y = "Debt Ratio")

ggplot(training_df, aes(x = factor(Bankrupt.), y = Net.Income.to.Total.Assets, fill = factor(Bankrupt.))) +
  geom_boxplot() +
  labs(x = "Bankrupt", y = "Net income to Total Assets")

ggplot(training_df, aes(x = factor(Bankrupt.), y = Net.worth.Assets, fill = factor(Bankrupt.))) +
  geom_boxplot() +
  labs(x = "Bankrupt", y = "Net worth Assets")

```

It seems that the probability of a company going bankrupt is quite small and a lot less frequent. We know that many variables are strongly correlated to each other so some methods that may be most suitable is dimensionality reduction like PCA or PLS for logistic regression, or tree based models that handle multicollinearity beter. 

